---
title: "Spectrum_methylation_analysis_TumNormal_WGBS_with_BSmooth_R"
author: "Samuel Terkper Ahuno"
date: "2023-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#samuel ahuno
#methylation analysis with RBSS with BSmooth r

### how to run the rmd as job script
1. run `knit` function
2. run `bsub` line
<!-- ## section x

#To run script
<!-- function knit() {
module load R/R-4.1.2
Rscript --max-ppsize=500000 -e "rmarkdown::render('$1', output_dir='$2')"
} -->
<!-- export -f knit

#set path to Rscripts
bsub -J "methyl_Bisulphite_spect" -R "rusage[mem=24]" -n 32 -o methyl_Bisulphite_spect.%J.out -e methyl_Bisulphite_spect.%J.err -W 5:00 "knit /juno/work/greenbaum/users/ahunos/methyl_SPECTRUM/scripts/BSmooth_WGBS_methylationAnalysisis_TumNormal.Rmd /juno/work/greenbaum/users/ahunos/methyl_SPECTRUM/scripts/" -->

```{r Loadlibraries, echo=FALSE}
library(bsseq)
library(magrittr)
library(data.table)
library(BiocParallel)#used for parallel analysis
```


## set paths
```{r setPaths}
#set paths to data and analysis directories
path_analysis_home <- "/juno/work/greenbaum/users/ahunos/methyl_SPECTRUM"
path_sepectrum_data <- "/work/greenbaum/projects/methylation_projects/SPECTRUM"

# tumPath = "/work/greenbaum/projects/methylation_projects/SPECTRUM/nf_core_methylseq_tum/OUTDIR/bismark/methylation_calls/methylation_coverage/SA123T_1_val_1_bismark_bt2_pe.deduplicated.bismark.cov.gz"
# normPath =  "/work/greenbaum/projects/methylation_projects/SPECTRUM/nf_core_methylseq/OUTDIR/bismark/methylation_calls/methylation_coverage/SA123_1_val_1_bismark_bt2_pe.deduplicated.bismark.cov.gz"

dir_2_create <- "bsmooth_methy"
# dir.create(file.path(dir_2_create))
# dir.create(file.path(dir_2_create, c("data")))
wrkDir <- paste0(getwd(),"/",dir_2_create)
```


## section; load files

```{r loadFiles}

##pull all coverage files 
files_cpG <- list.files(path_sepectrum_data, full.names = TRUE,
                        recursive = TRUE, pattern = "*.deduplicated.bismark.cov.gz") %>%
                            as_tibble()
names(files_cpG) <- "File"

#filter files of interest
files_cpG_cov <- files_cpG %>% 
                    dplyr::filter(str_detect(File, "nf_core_methylseq"))  %>% 
                        dplyr::filter(str_detect(File, "OUTDIR")) %>% 
                        mutate(Sample = basename(File),
                                Sample = case_when(str_detect(Sample, "SA123_") ~ "SA123N",
                                                 str_detect(Sample, "SA123T") ~ "SA123T"),
                                chemistry = "methylSeq",
                                treatment = case_when(str_detect(Sample, "SA123N") ~ "Normal",
                                                     str_detect(Sample, "SA123T") ~ "Tumor"),
                                Sample = gsub("_1.*","",Sample))  %>% 
                            as.data.frame()
#View(files_cpG_cov)
files_cpG_cov
```


create master file of files, samples, groups to run bsmooth

```{r metadata_EdgeR}
sample_metadata <- files_cpG_cov %>% mutate(Group = treatment)
sample_metadata
```



# Actual Differential methylations analysis
### Read bismark files from disk

```{r LoadBismark}
#dmr
#files <- sample_metadata$File
bismark_bsseq <- read.bismark(files = sample_metadata$File,
                    colData = DataFrame(row.names = sample_metadata$Sample, Type = sample_metadata$Group, Pair = c("Pair1", "Pair1")),
                      rmZeroCov = FALSE,
                      strandCollapse = FALSE,
                      verbose = TRUE, 
                      BACKEND = "HDF5Array",
                      BPPARAM = MulticoreParam(workers = 16)
                      )
bismark_bsseq

pData(bismark_bsseq)
#try combing samples to have replicates
bismark_bsseq_combine <- combine(bismark_bsseq, bismark_bsseq[,1])
```


```{r QC_1}
sampleNames(bismark_bsseq)
data(bismark_bsseq) #how does the dara look like

#check phenoData
pData(bismark_bsseq)

#see first few granges
head(granges(bismark_bsseq), n = 4)

```


```{r QC_2}
#check also have the M and Cov matrices
head(getCoverage(bismark_bsseq, type = "M"), n = 4)

#check coverage for all
head(getCoverage(bismark_bsseq), n = 4)

##data
dim(bismark_bsseq)
round(colMeans(getCoverage(bismark_bsseq)), 1)
# SA123T SA123N 
#    7.6    6.6 
#therefore average coverage per cpg of each sample is above
## Number of CpGs in all chromosome
length(bismark_bsseq)
#[1] 55,261,449

## Number of CpGs which are covered by at least 1 read in all 2 samples
sum(rowSums(getCoverage(bismark_bsseq) >= 1) == 2)
#[1] 53,947,003

## Number of CpGs with 0 coverage in all samples
sum(rowSums(getCoverage(bismark_bsseq)) == 0)
#[1] 0

#expectation of the CpGs to have at least one sample with zero coverage.
logp <- ppois(0, lambda = 7, lower.tail = FALSE, log.p = TRUE)
round(1 - exp(2 * logp), 3)
```


```{r}
#parallization
## Split data and run bs
# BS1 <- bismark_bsseq[, 1]
# saveRDS(BS1, file =paste0(wrkDir,"/SA123T.rds"))

# BS2 <- bismark_bsseq[, 2]
# saveRDS(BS2, file =paste0(wrkDir,"/SA123N.rds"))

```


### Run bsmoothing
```{r BSmooth}
#library(parallel)
bismark_bsseq.fit <- BSmooth(
    BSseq = bismark_bsseq, 
    BPPARAM = MulticoreParam(workers = 24), 
    verbose = TRUE)
```


```{r}

```


```{r}

```







## section; chromosomes of interset

```{r Compute_Tstats}
#filter out regions where 2 or more samples have less than 2x coverage
# BS.cov <- getCoverage(bismark_bsseq.fit)
# keepLoci.ex <- which(rowSums(BS.cov[, bismark_bsseq.fit$Type == "Tumor"] >= 2) >= 1 &
#                      rowSums(BS.cov[, bismark_bsseq.fit$Type == "Normal"] >= 2) >= 1)
# length(keepLoci.ex)

# BS.cancer.ex.fit <- BS.cancer.ex.fit[keepLoci.ex,]

bismark_bsseq.tstat <- BSmooth.tstat(bismark_bsseq.fit, 
                                    group1 = c("SA123T"),
                                    group2 = c("SA123N"), 
                                    estimate.var = "group2",
                                    local.correct = TRUE,
                                    verbose = TRUE)
bismark_bsseq.tstat

# $ find /work/greenbaum/TEMP -user ahunos | xargs rm -r * | head
# $ find /work/greenbaum/TEMP -uid 164079095 | xargs l | head

```

## section; plot differentailly methylated regions


```{r plotDMRs}
#plot DMR
pData <- pData(bismark_bsseq.fit)
pData$col <- rep(c("red", "blue"), each = 1)
pData(bismark_bsseq.fit) <- pData

```


<!-- ## section x

#To run script
<!-- function knit() {
module load R/R-4.1.2
Rscript --max-ppsize=500000 -e "rmarkdown::render('$1', output_dir='$2')"
} -->
<!-- export -f knit

#set path to Rscripts
rmd_path=/juno/work/greenbaum/users/ahunos/methyl_SPECTRUM/scripts/methylation_Bisulphite_Spectrum.Rmd 

bsub -J "methyl_Bisulphite_spect" -R "rusage[mem=24]" -n 32 -o methyl_Bisulphite_spect.%J.out -e methyl_Bisulphite_spect.%J.err -W 5:00 "knit /juno/work/greenbaum/users/ahunos/methyl_SPECTRUM/scripts/methylation_Bisulphite_Spectrum.Rmd /juno/work/greenbaum/users/ahunos/methyl_SPECTRUM/scripts/" -->
